The promise of community-driven preprints in ecology and evolution

### 2024-05-04 ms/ms.qmd
Fixing up last few sections

Publishing preprints -- papers communicating non-peer-reviewed research findings -- is now an entrenched practice across a multitude of scientific disciplines [@Ginsparg2011]. While preprint use in biology has had a slower uptake [@Berg2016], in ecology and evolution, new discipline-specific preprint servers, such as *EcoEvoRxiv*, provide a means by which ecologists and evolutionary biologists can disseminate research findings. Preprints are expected to promote open science, increase the visibility of research (e.g., [@fu2019releasing]), and facilitate the rapid sharing of new discoveries that can have significant impacts on fundamental and applied knowledge globally. Preprints attempt to break down barriers to scientific publishing by: 1) increasing the speed, visibility and output of research findings; 2) helping establish precedence of research findings; 3) removing financial barriers to publication; and 4) encouraging constructive and critical feedback from the research community [@Proulx2013; @bourne2017ten; @vale2015accelerating].

Preprint servers can empower researchers to make their findings more accessible, open, and transparent but only if they are used as forums for spreading and discussing findings within a research community. Significant barriers to preprint use still exist globally – from a lack of clarity around preprint policies at journals [@Klebel2020] to a stigma that preprints are viewed by the research community to be of poor quality [@Chiarelli2019] (but see [@Faser2022]). Nonetheless, we lack a systematic understanding of the factors that influence preprint use in ecology and evolution that can help inform future initiatives and allow us to work harder in further breaking down barriers to preprint use in our field.

*EcoEvoRxiv* is one of the few community-driven preprint servers that has paved the way for new initiatives, including by accepting multilingual preprints, recognising registered reports and non-traditional research reports, and promoting peer review and community discussion to improve the quality of preprints and speed up their peer-reviewed publication (e.g., Peer Community In -- PCI [@guillemaud2019peer]). Peer Community In allows for fast constructive peer review around a preprint. Peer reviews are transparent and published online and the community makes a recommendation on where the reasearch findings should be published [@guillemaud2019peer].

Here, we explore the first preprints uploaded to *EcoEvoRxiv* to understand preprint practices and uses in ecology and evolution. We focus our attention on better understanding: 1) what countries authors who use *EcoEvoRxiv* are located; 2) how career stage and gender impact preprint server use; 3) the extent to which authors make use of preprint servers for registered reports and community-driven peer review; 4) the extent to which data and code are shared in preprints; and 5) how many preprints remain unpublished, and for those that are published, how long it took them to become published. In the process, we also provide a summary of what makes *EcoEvoRxiv* distinct from other preprint servers to help further educate researchers about the benefits of using community-driven preprint servers to post their research findings.

@@

*EcoEvoRxiv* is run by the Society for Open, Reliable, and Transparent Ecology and Evolutionary biology (SORTEE)[@o2021towards]. Originally launched in 2018 on the Center for Open Science preprint platform, *EcoEvoRxiv* has grown to become a popular preprint server for researchers in ecology and evolutionary biology. It has since been adopted by the California Digital Library (CDL) to be hosted in perpetuity without fees. Editors are researchers from across the globe (currently eight different countries) in the fields of ecology and evolutionary biology who volunteer their time to screen papers and push new initiatives in the preprint space. *EcoEvoRxiv* allows authors to post both preprints and postprints, or author-accepted manuscripts -  a version of an accepted or published article without typesetting and formatting by a journal. Postprints are a useful way to ensure published articles are openly accessible and can be posted anytime after publication (assuming the journal has a preprint policy that recognises preprints-- though most do; see https://www.sherpa.ac.uk/romeo/). To better understand preprint (and postprint) use on *EcoEvoRxiv*, we downloaded metadata on the articles available on *EcoEvoRxiv* as of `r gsub("UTC", "", dates[2])` (see Supplement for more details on methods). For the purpose of summarising articles on *EcoEvoRxiv* we include both preprints and postprints as 'articles' in our overview analysis.

@@

### *Cautious 'open'-mindedness of research in preprints*
In addition to speeding up dissemination, preprints can also be a useful way to ensure that research remains open and accessible to the research community because the paper is available online no matter what the open access status of the publication ends up being [@bourne2017ten; @vale2015accelerating]. Using the published articles on *EcoEvoRxiv*, we evaluated the open access status of each paper using the R package *roadoi* to connect to Unpaywall platform [@jahn2024]. Most of the published preprints ended up being open access ([@fig-pubsummary]A), however, the status of many `r sum_is_oa$n[3]` (`r sum_is_oa$percent[3]*100`%) were unknown, and `r sum_is_oa$n[1]` (`r sum_is_oa$percent[1]*100`%) were not published in open access research journals. The level of open access for a paper published in an open-access journal also varied widely ([@fig-pubsummary]B).

Data and code sharing are also key components of open science [@roche2015public]. In the spirit of 'openness' we expected data and code sharing among preprints to be better than in many papers published in research journals [@roche2015public; @o2021preferred]. Despite this, we found that `r preprint_data[1,2]` (`r preprint_data[1,"percent"]*100`%) of relevant preprints on *EcoEvoRxiv* (i.e., only where data was generated) did not share data, while `r preprint_code[1,2]` (`r preprint_code[1,"percent"]*100`%) did not share code. Authors may be reluctant to share data and code for preprints because of the perceived concern that others may acquire and use such data and code without authors' consent. However, while `r pre_article_data[3,2]` (`r pre_article_data[3,"percent"]*100`%) of preprints that did not share data at the preprint stage did ultimately share data in the published article, there were still `r pre_article_data[2,2]` (`r pre_article_data[2,"percent"]*100`%) that never shared data. The same was true for code. Overall, `r pre_article_code[3,2]` (`r pre_article_code[3,"percent"]*100`%) of preprints did not share code at the preprint stage but did in the published article; however, `r pre_article_code[2,2]` (`r pre_article_code[2,"percent"]*100`%) never shared code. Code and data- sharing patterns in our sample of preprints are consistent with analyses of published articles (e.g., [@o2021preferred]), even for journals with strict public data archiving policies [@roche2015public].

@@

Our data-driven approach has allowed us to better understand some of the key elements of uncertainty surrounding preprints and identify shortcomings and areas needing greater clarity to authors. Several key insights follow from our analysis of the first *EcoEvoRxiv* preprints.

- First, authors need not be afraid of sharing data and code early in the preprint stage. Sharing data and code early can help improve the quality of research, establish precedence, and improve the transparency and computational reproducibility of scientific findings [@sarabipour2019value]. If authors are worried about data being used unintentionally, clear information surrounding its reuse and withholding some data or meta-data needed to understand the dataset may be options. 
- Second, authors need to be aware of the benefits of posting preprints and post-prints for constructive feedback [@bourne2017ten]. While it is clear preprints help establish precedence and allow findings to be openly accessible, it is still rare that constructive discussions form around preprints in an open form, which is consistent with findings on other preprint servers (e.g. bioRxiv [@anderson2020biorxiv]). Clearly, as a community, we need to provide better platforms that document discussions around preprint findings. Such discussions help the authors improve their work and communicate their findings more effectively (when done constructively, of course). One way to facilitate such discussions may be to use open peer reviewing services such as *Peer Community In* (PCI) or *PubPeer* to provide feedback on preprints that ultimately is associated with a preprint.
- Third, it remains clear that the time between preprinting and publication is still quite long. One possible explanation is that preprints are not being sent to suitable journals. Interestingly, out of the `r nrow(data2)` preprints, only `r pci[2,2]` (`r pci[2,3]*100`%) used Peer Community In (PCI) as a preprint peer-review service. Using PCI provides authors with quicker reviews of a paper and a more realistic recommendation of where such work is likely to be published. Many journals now also accept PCI reviews and recommendations when publishing a paper. As such, we would encourage more authors to consider sending their preprints for PCI review to possibly speed up the publication process. 
- Finally, while most preprints get seamlessly connected with and merged with their published version some remain disconnected. Incorrect cross-linking can create confusion and lead to frustration among authors. Thankfully, the reasons for unmatched preprints and publications are well understood and easily rectified. They often result from a mismatch between preprint and published metadata (e.g., titles and author details). In fact, it was reasonably common for titles of preprints to change between the preprint and the published article (`r titles[2,2]` or `r titles[2,3]*100`%). We found that mismatched metadata almost always contributed to preprints and published articles not being matched automatically in Google Scholar. We recommend that authors ensure that, when preprints are published in a journal, they update their preprint with the publication DOI. Especially if their title has changed, this is quite easy for authors to do on *EcoEvoRxiv*, and it will ensure that the preprint is linked to the published article and that the citations are merged. It will also create a link between the final preprint and the published paper so that, no matter the open access status of the publication, the research findings will remain openly accessible, saving costs for authors and improving the visibility and use of research.

@@

# Acknowledgements
We would like to thank the California Digital Library (CDL) and the CDL team (particularly Alainna Wrigley) for their support in hosting and maintaining *EcoEvoRxiv* for the Society for Open, Reliable, and Transparent Ecology and Evolution (SORTEE). We would also like to thank Gabriela Hidalgo and Daisy Larios for helping connect us with the IUCN and facilitating discussions to make *EcoEvoRxiv* a place where IUCN reports can be posted. Finally, we would also like to thank the endless number of SORTEE volunteers, and those especially on the *EcoEvoRxiv* Committee, who have helped to make *EcoEvoRxiv* a success. This paper emerged from a hackathon at the 2023 SORTEE conference, and we thank delegates who attended the session but could not be part of this paper. DWAN would also like to thank the Australian Research Council for a Future Fellowship (FT220100276).

>>>>

Publishing preprints -- papers communicating non-peer-reviewed research findings -- is now an entrenched practice across a multitude of scientific disciplines [@Ginsparg2011]. While preprint use in biology has had a slower uptake [@Berg2016], in ecology and evolution, new discipline-specific preprint servers, such as *EcoEvoRxiv*, provide a means by which ecologists and evolutionary biologists can disseminate research findings. Preprints are expected to promote open science, increase the visibility of research (e.g., [@fu2019releasing]), and facilitate the rapid sharing of new discoveries that can have significant impacts on fundamental and applied knowledge globally. Preprints attempt to break down barriers to scientific publishing by: 1) increasing the speed, visibility and output of research findings which can lead to more citations [e.g., @colavizza2024analysis]; 2) helping establish precedence of research findings; 3) removing financial barriers to publication; and 4) encouraging constructive and critical feedback from the research community [@Proulx2013; @bourne2017ten; @vale2015accelerating].

Preprint servers can empower researchers to make their findings more accessible, open, and transparent but only if they are used as forums for spreading and discussing findings within a research community. Significant barriers to preprint use still exist globally – from a lack of clarity around preprint policies at journals [@Klebel2020] to a stigma that preprints are viewed by the research community to be of poor quality [@Chiarelli2019] (but see [@Faser2022]). Nonetheless, we lack a systematic understanding of the factors that influence preprint use in ecology and evolution. Such an understanding can help inform future initiatives and allow us to work harder in further breaking down barriers to scientific publishing.

*EcoEvoRxiv* is one of the few community-driven preprint servers that has paved the way for new initiatives, by accepting multilingual preprints, recognising registered reports and non-traditional research reports, and promoting peer review and community discussion which we hope will improve the quality of preprints and speed up their peer-reviewed publication (e.g., Peer Community In -- PCI [@guillemaud2019peer]). Peer Community In allows for fast constructive peer review around a preprint. Peer reviews are transparent and published online and the community makes a recommendation on where the reasearch findings should be published [@guillemaud2019peer].

Here, we explore the first 1000+ preprints uploaded to *EcoEvoRxiv* to understand preprint practices and uses in ecology and evolution. We focus our attention on better understanding: 1) in what countries authors who use *EcoEvoRxiv* are located; 2) how career stage and gender impact preprint server use; 3) the extent to which authors make use of preprint servers for registered reports and community-driven peer review; 4) the extent to which data and code are shared in preprints; and 5) how many preprints remain unpublished, and for those that are published, how long it took them to become published. In the process, we also provide a summary of what makes *EcoEvoRxiv* distinct from other preprint servers to help further educate researchers about the benefits of using community-driven preprint servers to post their research findings.

@@

*EcoEvoRxiv* is run by the Society for Open, Reliable, and Transparent Ecology and Evolutionary biology (SORTEE)[@o2021towards]. Originally launched in 2018 on the Center for Open Science preprint platform, *EcoEvoRxiv* has grown to become a popular preprint server for researchers in ecology and evolutionary biology. It has since been adopted by the California Digital Library (CDL) to be hosted in perpetuity without fees. Editors are researchers from across the globe (currently eight different countries) in the fields of ecology and evolutionary biology who volunteer their time to screen papers and push new initiatives in the preprint space. *EcoEvoRxiv* allows authors to post both preprints and postprints, or author-accepted manuscripts -  a version of an accepted or published article without typesetting and formatting by a journal. Postprints are a useful way to ensure published articles are openly accessible and can be posted anytime after publication (assuming the journal has a policy that recognises pre- and post-prints -- though most do; see https://www.sherpa.ac.uk/romeo/). To better understand preprint (and postprint) use on *EcoEvoRxiv*, we downloaded metadata on the articles available on *EcoEvoRxiv* as of `r gsub("UTC", "", dates[2])` (see Supplement for more details on methods). For the purpose of summarising articles on *EcoEvoRxiv* we include both preprints and postprints as 'articles' in our overview analysis.

@@

### *Cautious 'open'-mindedness of research in preprints*
In addition to speeding up dissemination, preprints can also be a useful way to ensure that research remains open and accessible to the research community because the paper is available online no matter what the open access status of the publication ends up being [@bourne2017ten; @vale2015accelerating]. Using the published articles on *EcoEvoRxiv*, we evaluated the open access status of each paper using the R package *roadoi* (vers. `r utils::packageVersion("roadoi")`) to connect to the Unpaywall platform [@jahn2024]. Most of the published articles arising from preprints were open access [`r sum_is_oa$n[2]` (`r sum_is_oa$percent[2]*100`%) [@fig-pubsummary]A]; however, the status of many were unknown [`r sum_is_oa$n[3]` (`r sum_is_oa$percent[3]*100`%)], and `r sum_is_oa$n[1]` (`r sum_is_oa$percent[1]*100`%) were published behind paywalls. The type of open access for a published paper in an open-access journal also varied widely ([@fig-pubsummary]B).

Data and code sharing are also key components of open science [@roche2015public]. In the spirit of 'openness', we expected data and code sharing among preprints to be greater than in many papers published in research journals [@roche2015public; @o2021preferred]. Despite this, we found that `r preprint_data[1,2]` (`r preprint_data[1,"percent"]*100`%) of preprints relying on data on *EcoEvoRxiv* did not share data, and `r preprint_code[1,2]` (`r preprint_code[1,"percent"]*100`%) did not share code. Authors may be reluctant to share data and code for preprints because of the perceived concern that others may acquire and use such data and code prior to publication in a journal. Notably, authors of `r pre_article_data[3,2]` (`r pre_article_data[3,"percent"]*100`%) preprints that did not share data at the preprint stage ultimately did share data when the article was published; authors for `r pre_article_data[2,2]` (`r pre_article_data[2,"percent"]*100`%) preprints never shared data. The same was true for code. Overall, `r pre_article_code[3,2]` (`r pre_article_code[3,"percent"]*100`%) preprints had no open code at the preprint stage but did at the published article stage; authors of `r pre_article_code[2,2]` (`r pre_article_code[2,"percent"]*100`%) preprints did not share code on publication. Code and data-sharing patterns in our sample of preprints are consistent with analyses sharing practices for published articles (e.g., [@o2021preferred]), even for journals with strict public data archiving policies [@roche2015public].

@@

Our data-driven approach has allowed us to better understand some of the key elements of uncertainty surrounding preprints and identify shortcomings and areas in need of clarification for authors. Several key insights follow from our analysis of the first *EcoEvoRxiv* preprints:

- First, authors are encouraged to share their data and code at the preprint stage. Sharing data and code early can help improve the quality of research, establish precedence, and improve the transparency and computational reproducibility of scientific findings [@sarabipour2019value]. Reassuringly, sharing data and code is rarely associated with 'scooping' of research findings [@soeharjono2021reported]. If authors are worried about data being used unintentionally, clear information surrounding its reuse can include in a license (see https://choosealicense.com). Data can also be achived with an embargo on its reuse [@roche2014troubleshooting]. 
- Second, authors need to be aware of the benefits of posting preprints and postprints for constructive feedback [@bourne2017ten]. While it is clear preprints help establish precedence and allow findings to be openly accessible, it is still seems rare that constructive discussions form around preprints in an open forum (e.g. bioRxiv [@anderson2020biorxiv]). Unfortunately, the *EcoEvoRxiv* website does not provide opportunities for discussion given the limitations of the web server at this point in time. As such, we could not accurately assess how much discourse around a given preprint occurs. Clearly, as a community, we need to provide better platforms that document discussions around preprint findings. Such discussions help the authors improve their work and communicate their findings more effectively (when done constructively, of course). One way to facilitate such discussions may be to use open peer reviewing services, such as *Peer Community In* (PCI) or *PubPeer* (see also [@avissar2024recommendations] for more options), to provide feedback on preprints.
- Third, it remains clear that the time between posting a preprint and publication is still quite long. One possible explanation is that preprints are not being sent to suitable journals. Interestingly, out of the `r nrow(data2)` preprints, only `r pci[2,2]` (`r pci[2,3]*100`%) used Peer Community In (PCI) as a preprint peer-review service. Using PCI provides authors with quicker reviews of a paper and a more realistic recommendation of where such work is likely to be published. Many journals now also accept PCI reviews and recommendations when publishing a paper. As such, we would encourage more authors to consider sending their preprints for PCI review to possibly speed up the publication process. 
- Finally, while most preprints get seamlessly connected with and merged with their published version some remain disconnected. Incorrect cross-linking can create confusion and lead to frustration among authors. Thankfully, the reasons for unmatched preprints and publications are well understood and easily rectified. They often result from a mismatch between preprint and published metadata (e.g., titles and author details). In fact, it was reasonably common for titles of preprints to change between the preprint and the published article (`r titles[2,2]` or `r titles[2,3]*100`%). We found that mismatched metadata almost always contributed to preprints and published articles not being matched automatically in Google Scholar. We recommend that authors ensure that, when preprints are published in a journal, they update their preprint with the publication DOI. Especially if their title has changed, this is quite easy for authors to do on *EcoEvoRxiv*, and it will ensure that the preprint is linked to the published article and that the citations are merged. It will also create a link between the final preprint and the published paper so that, no matter the open access status of the publication, the research findings will remain openly accessible, saving costs for authors and improving the visibility and use of research.

@@

# Acknowledgements
We would like to thank the California Digital Library (CDL) and the CDL team (particularly Alainna Wrigley) for their support in hosting and maintaining *EcoEvoRxiv* for the Society for Open, Reliable, and Transparent Ecology and Evolutionary biology (SORTEE). We would also like to thank Gabriela Hidalgo and Daisy Larios for helping connect us with the IUCN and facilitating discussions to make *EcoEvoRxiv* a place where IUCN reports can be posted. Finally, we would also like to thank the endless number of SORTEE volunteers, and those especially on the *EcoEvoRxiv* Committee, who have helped to make *EcoEvoRxiv* a success. This paper emerged from a hackathon at the 2023 SORTEE conference, and we thank delegates who attended the session but could not be part of this paper. DWAN would also like to thank the Australian Research Council for a Future Fellowship (FT220100276).
